---
title: "queeems: Quantify the Extent of Evolutionary Evidence in Molecular
Sequences"
author:
  - name: Hassan Sadiq
    affiliation:
    - Department of Statistics and Actuarial Science,
        Stellenbosch University, South Africa
    email: hassan.t.sadiq@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
package: queeems
bibliography: queeems.bib
vignette: |
    %\VignetteIndexEntry{queeems Tutorial}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
    BiocStyle::markdown()
```

# Introduction
As with every other statistical investigation, accuracy of inferences from
analyses of genetic sequences depends on the data quality. Assessing the
quality of molecular sequences for biological studies requires quantifying
the degree of evolutionary signature retained in the data. Such an assessment
is popularly done using the saturation (*H*) index introduced by @xia2002.
The index is an extension of @shannon1948 information entropy and it is not
robust to the evolutionary tree that must have generated the data. In general,
a single H index is estimated for an alignment of protein sequences. The
significance of the evolutionary information retained in the data is then
assessed with the Student's t hypothesis test that compares the estimated
index to a pre-determined threshold. Xia's H index is an invaluable
contribution to computational evolutionary biology studies. It is however
best suited for phylogeny reconstruction inquiries. `queeems` is a R
[@cran2025] package hosted on the Bioconductor [@bioc2004] platform that
contains Bayesian functions that can be used to assess the quality of
protein sequences beyond just for phylogeny reconstruction purposes.
It can also be used to assess how suitable protein sequences are for broader
evolutionary analyses, including natural selection investigations. In this
practical application of some of the primary functions in the package, I use
simulated genetic sequences to demonstrate some utilities of the package.


# Installation
Use the following code to install `queeems` from the Bioconductor platform.
```{r, eval=FALSE}
if(!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("queeems")
```
The developmental version of the package that contains the most recent
updates, is available under the `devel` tab of the Bioconductor platform
and can be installed by running the following code, just before the final
install line above: `BiocManager::install(version="devel")`.

# Data
The molecular sequences analysed herein are generated with the `scoup`
[@sadiq2024] package that allows genetic sequences to be simulated following
a simplified Ornstein-Uhlenbeck (OU) process. A R file that contains the
simulation code and a folder that contains the simulated data are available
in the `vignettes/data/` folder as part of the `queeems` package. Results from
analyses of an empirical `VertCOI.fasta` data from DAMBE [@xia2018] is also
presented below. The data can also be accessed in the `inst/extdata/` folder
of the `queeems` package.

# Simulation Studies
Genetic sequence alignments were generated on a balanced bifurcating
evolutionary tree that has 64 leaves. For each data simulated, the length
of the internal and the terminal branches were all kept equal to `0.10` for
each execution. The *variance of synonymous selection coefficients* was
specified as, $\sigma_{s}^{2}=0.01$ to ensure that positive natural selection
effect [@spielman2015] will be detected. Eight data replicates were
generated. Other necessary `scoup` simulation parameters specified include:
*number of codon sites = 250*, *OU reversion rate = 0.00*,
*OU asymptotic variance = 0.00*, *OU asymptotic mean = 0.00*, *variance of
non-synonymous selection coefficients = 0.50*. The data analysis was
initiated with the following code.  

```{r, eval=TRUE}
# ><>< # Load package into current R session
library("queeems")

# ><>< # Retrieve path to simulated tutorial data folder
datadir <- system.file("extdata/example/", package="queeems")

# ><>< # Read merged molecular sequences data as text
seqsname <- file.path(datadir, "tut_seqs.txt")
seqstext <- readLines(seqsname)

# ><>< # Extract information about number of simulated replicates
sbreaks <- which(seqstext == "  64  750")
nreps <- length(sbreaks)

# ><>< # Create data frames to save interesting outputs
homoBF <- vector("numeric", nreps)
cnames <- paste0("r", seq(1,nreps))
nucent <- matrix(NA, 750, nreps, dimnames=list(NULL,cnames))
codent <- matrix(NA, 250, nreps, dimnames=list(NULL,cnames))
aasent <- matrix(NA, 250, nreps, dimnames=list(NULL,cnames))
sbayes <- matrix(NA, 250, nreps, dimnames=list(NULL,cnames))

# ><>< # Storage for saturation-free data to allow subsequent analyses
nosatpath <- file.path(tempdir(), "satfree.txt")

# ><>< # Adhoc function to extract saturation-free data
printsite <- function(b0, sqarray, ipath=nosatpath){
    ntag <- rownames(sqarray)[b0]
    sqline <- paste(sqarray[b0,], collapse="")
    fulltxt <- paste0(">", ntag, "\n", sqline)
    write.table(fulltxt, ipath, TRUE, FALSE, col.name=FALSE, row.names=FALSE)
    return(0)
}

# ><>< # Analyse data replicates captured from merged data file
for(a0 in seq(1,nreps)){

    # ><>< # Extract and temporarily save a sequence replicate
    init <- sbreaks[a0] + 1
    halt <- sbreaks[a0] + 128    # >< # Each replicate is from 64 leaves
    seqrep <- seqstext[seq(init,halt)]
    tempfile <- file.path(tempdir(), "iseqs.fasta")
    write.table(seqrep, file=tempfile, append=FALSE,
        quote=FALSE, row.names=FALSE, col.names=FALSE)

    # ><>< # Execute saturation analysis
    satest <- seqSaturation(tempfile, "A", 6, 0.4, 7)
    homoBF[a0] <- as.numeric( summary(satest)["fullBF",])
    sbayes[,a0] <- BFs(satest)

    # ><>< # Extract and save saturation-free sequence locations
    nullsites <- which(BFs(satest) < 100)
    sieved <- seqfilter(tempfile, nullsites, "codon")
    imatrix <- as.matrix(sieved)
    itext <- paste0("  ", nrow(imatrix), "  ", ncol(imatrix))
    write.table(itext,nosatpath,a0!=1,FALSE,row.names=FALSE,col.names=FALSE)
    pholder <- vapply(seq(1,nrow(imatrix)), printsite, imatrix, FUN.VALUE=0)
    write.table("\n",nosatpath,TRUE,FALSE,row.names=FALSE,col.names=FALSE)

    # ><>< # Execute codon entropy analysis
    entcdn <- molentropy(tempfile, "codon", "shannon")
    codent[,a0] <- sitentropies(entcdn)

    # ><>< # Process nucleotide entropy analysis
    entnuc <- molentropy(tempfile, "dna", "shannon")
    nucent[,a0] <- sitentropies(entnuc)

    # ><>< # Transform nucleotide to amino acid sequences
    dnaformat <- Biostrings::readDNAStringSet(tempfile)
    aaformat <- Biostrings::translate(dnaformat)
    Biostrings::writeXStringSet(aaformat, tempfile)

    # ><>< # Implement amino acid entropy analysis
    entamino <- molentropy(tempfile, "aa", "shannon")
    aasent[,a0] <- sitentropies(entamino)

    # ><>< # Provide analysis progress update
    message("Completed analyses of data replicate ", sprintf("%02.0f",a0))
}
```

Analytical estimates of the ratio of non-synonymous to synonymous substitution
rates ($\mathrm{d}N/\mathrm{d}S$) are obtained from `scoup` where they are
calculated using expressions from @spielman2015. The maximum likelihood
estimates of the ratio ($\omega$) are obtained using PAML [@yang2007] and
the control files used for the execution of the analyses are available in
the `vignettes/data/` folder in the `queeems` package. The correlation
between the entropy estimates and the Bayes factors demonstrate consistency
of the overall inferences with the entropy-based technique used in DAMBE
[@xia2018].

# Empirical Analyses
The `VertCOI.fasta` data [@xia2018] contains mitochondrial COI sequences from
eight vertebrates including: *Masturus lanceolatus* (sunfish), *Homo sapiens*
(human), *Bos taurus* (cow), *Balaenotera musculus* (blue whale), *Pongo
pygmaeus* (Bornean orangutan), *Pantroglodytes* (chimpanzee), *Gallus gallus*
(chicken) and *Alligator mississippiensis* (American alligator) [@xia2009].
Each sequence comprise 504 codon bases. The results obtained from analysing
the data using DAMBE [@xia2018] (output screenshots are also available in
the `vignettes/data/` folder) and `queeems` are summarised below.

```{r, eval=TRUE}
# Analyse data with queeems
# Generate nice summary plots
```

---
|DAMBE                |        |queeems              |        |
|:--------------------|:------:|:--------------------|:------:|
|Invariant sites prop.|0.0000  |Invariant sites prop.|a       |
|Mean H               |0.4457  |Saturated sires      |a       |
|Std. error of H      |0.0148  |Overall Bayes factor |a       |
|Iss                  |0.2698  |Overall decision     |a       |
|Iss.c (symmetric)    |0.8093  |Overall decision     |Not sat.|
|Iss.c (asymmetric)   |0.7095  |                     |        |
|Overall decision     |Not sat.|                     |        |
Table: Comparative saturation analyses outputs that demonstrate the
effectiveness of the inference functions in `queeems`.

Inferences from DAMBE show that the data is not saturated given that, for
both the symmetrical and asymmetrical trees respectively, the critical
saturation index (`Iss.c = 0.8093 & 0.7095`) is greater than the saturation
index (`Iss = 0.2698`) estimated for the data. This inference is the same as
the overall inference obtained with the Bayesian approach with `queeems`. In
addition, `queems` made it possible to identify sites `...`as saturated sites.
In fact, excluding the identified sites `increased/reduced` the value of the
homogeneous maximum likelihood estimate of the ratio of non-synonymous
to synonymous substitution rate ($\omega$) from ... to .....

Entropy estimates are not directly used as the primary saturation inference
measures in `queeems`, unlike in DAMBE [@xia2018]. There are however functions
in `queeems` that can be used to obtain different entropy estimates with
respect to molecular data. For example, Figure 1 below illustrates ....
```{r, eval=TRUE}
# Analyse data with queeems
# Generate nice summary plots
```

# Conclusion
I demonstrated that, compared to the popular existing software for analysing
molecular saturation, `queeems` avails the opportunity to allow researchers
more easily identify the codon sites contributing to overall protein sequence
saturation. I further showed that such knowledge may lead to improved estimate
of the magnitude of the natural selection pressure obtained from such protein
data. The choice of the weight parameters (`size = 6`, `proportion < 1 =
0.4`, `maximum = 7`) used during the `queeems` analyses presented here is
arbitrary. A study about the sensitivity of the parameters to the inferences
obtained is ongoing and intended to be submitted for publication upon
completion.

# Citation
A more appropriate citation will be provided for the package in due course.
In the meantime, kindly cite the package as, <em>Sadiq, H. 2026. "queeems:
Quantify the Extent of Evolutionary Evidence in Molecular Sequences".
Bioconductor R Package.</em>

# References

<div id="refs"></div>

# Session info

The output of `sessionInfo()` from the computer where this file is generated
is provided below.

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
