---
title: "queeems: Quantify the Extent of Evolutionary Evidence in Molecular
Sequences"
author:
  - name: Hassan Sadiq
    affiliation:
    - Department of Statistics and Actuarial Science,
        Stellenbosch University, South Africa
    email: hassan.t.sadiq@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
package: queeems
bibliography: queeems.bib
vignette: |
    %\VignetteIndexEntry{queeems Tutorial}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
    BiocStyle::markdown()
```

# Introduction
As with every other statistical investigation, accuracy of inferences from
analyses of genetic sequences depends on the data quality. Assessing the
quality of molecular sequences for biological studies requires quantifying
the degree of evolutionary signature retained in the data. Such an assessment
is popularly done using the saturation (*H*) index introduced by @xia2002.
The index is an extension of @shannon1948 information entropy and it is not
always robust to the evolutionary tree that must have generated the data. In
general, a single H index is estimated for an alignment of protein sequences.
The significance of the evolutionary information retained in the data is then
assessed with the Student's t hypothesis test that compares the estimated
index to a pre-determined threshold. Xia's H index is an invaluable
contribution to computational evolutionary biology studies. It is however
best suited for phylogeny reconstruction inquiries. `queeems` is a R
[@cran2025] package hosted on the Bioconductor [@bioc2004] platform that
contains Bayesian functions that can be used to assess the quality of
protein sequences beyond just for phylogeny reconstruction purposes.
It can also be used to assess how suitable protein sequences are for broader
evolutionary analyses, including natural selection investigations. In this
practical application of some of the primary functions in the package, I use
simulated genetic sequences to demonstrate some utilities of the package.


# Installation
Use the following code to install `queeems` from the Bioconductor platform.
```{r, eval=FALSE}
if(!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("queeems")
```
The developmental version of the package that contains the most recent
updates, is available under the `devel` tab of the Bioconductor platform
and can be installed by running the following code, just before the final
install line above: `BiocManager::install(version="devel")`.

# Data
The molecular sequences analysed herein are generated with the `scoup`
[@sadiq2024] package that allows genetic sequences to be simulated following
a simplified Ornstein-Uhlenbeck (OU) process. A R file that contains the
simulation code and a folder that contains the simulated data are available
in the `inst/extdata/example/` folder as part of the `queeems` package.

# Simulation Studies
Genetic sequence alignments were generated on a balanced bifurcating
evolutionary tree that has 64 leaves. For each data simulated, the length
of the internal and the terminal branches were all kept equal to `0.10` for
each execution. The *variance of synonymous selection coefficients* was
specified as, $\sigma_{s}^{2}=0.01$ to ensure that positive natural selection
effect [@spielman2015] will be detected. Eight data replicates were
generated. Other necessary `scoup` simulation parameters specified include:
*number of codon sites = 250*, *OU reversion rate = 0.00*,
*OU asymptotic variance = 0.00*, *OU asymptotic mean = 0.00*, *variance of
non-synonymous selection coefficients = 0.50*. The saturation analysis of
the simulated sequences was initiated with the following code that includes
storage of necessary outputs.  

```{r, eval=FALSE}
# ><>< # Load package into current R session
library("queeems")

# ><>< # Retrieve path to simulated tutorial data folder
datadir <- system.file("extdata/example", package="queeems")

# ><>< # Read merged molecular sequences data as text
seqsname <- file.path(datadir, "tut_seqs.txt")
seqstext <- readLines(seqsname)

# ><>< # Extract information about number of simulated replicates
sbreaks <- which(seqstext == "  64  750")
nreps <- length(sbreaks)

# ><>< # Create data frames to save interesting outputs
homoBF <- vector("numeric", nreps)
cnames <- paste0("r", seq(1,nreps))
nucent <- matrix(NA, 750, nreps, dimnames=list(NULL,cnames))
codent <- matrix(NA, 250, nreps, dimnames=list(NULL,cnames))
aasent <- matrix(NA, 250, nreps, dimnames=list(NULL,cnames))
sbayes <- matrix(NA, 250, nreps, dimnames=list(NULL,cnames))

# ><>< # Storage for saturation-free data to allow subsequent analyses
nosatpath <- file.path(tempdir(), "satfree.txt")

# ><>< # Adhoc function to extract saturation-free data
printsite <- function(b0, sqarray, ipath=nosatpath){
    ntag <- rownames(sqarray)[b0]
    sqline <- paste(sqarray[b0,], collapse="")
    fulltxt <- paste0(">", ntag, "\n", sqline)
    write.table(fulltxt, ipath, TRUE, FALSE, col.name=FALSE, row.names=FALSE)
    return(0)
}

# ><>< # Analyse data replicates captured from merged data file
for(a0 in seq(1,nreps)){

    # ><>< # Extract and temporarily save a sequence replicate
    init <- sbreaks[a0] + 1
    halt <- sbreaks[a0] + 128    # >< # Each replicate is from 64 leaves
    seqrep <- seqstext[seq(init,halt)]
    tempfile <- file.path(tempdir(), "iseqs.fasta")
    write.table(seqrep, file=tempfile, append=FALSE,
        quote=FALSE, row.names=FALSE, col.names=FALSE)

    # ><>< # Execute saturation analysis
    satest <- seqSaturation(tempfile, "A", 6, 0.4, 7)
    homoBF[a0] <- as.numeric( summary(satest)["fullBF",])
    sbayes[,a0] <- BFs(satest)

    # ><>< # Extract and save saturation-free sequence locations
    nullsites <- which(BFs(satest) < 100)
    sieved <- seqfilter(tempfile, nullsites, "codon")
    imatrix <- as.matrix(sieved)
    itext <- paste0("  ", nrow(imatrix), "  ", ncol(imatrix))
    write.table(itext,nosatpath,a0!=1,FALSE,row.names=FALSE,col.names=FALSE)
    pholder <- vapply(seq(1,nrow(imatrix)), printsite, imatrix, FUN.VALUE=0)
    write.table("\n",nosatpath,TRUE,FALSE,row.names=FALSE,col.names=FALSE)

    # ><>< # Execute codon entropy analysis
    entcdn <- molentropy(tempfile, "codon", "shannon")
    codent[,a0] <- sitentropies(entcdn)

    # ><>< # Process nucleotide entropy analysis
    entnuc <- molentropy(tempfile, "dna", "shannon")
    nucent[,a0] <- sitentropies(entnuc)

    # ><>< # Transform nucleotide to amino acid sequences
    dnaformat <- Biostrings::readDNAStringSet(tempfile)
    aaformat <- Biostrings::translate(dnaformat)
    Biostrings::writeXStringSet(aaformat, tempfile)

    # ><>< # Implement amino acid entropy analysis
    entamino <- molentropy(tempfile, "aa", "shannon")
    aasent[,a0] <- sitentropies(entamino)

    # ><>< # Provide analysis progress update
    message("Completed analyses of data replicate ", a0)
}
```

A few statistics necessary to assess the accuracy of the outputs from
`queeems` had to be obtained using other packages and applications. For
example, analytical estimates of the ratio of non-synonymous to synonymous
substitution rates ($\mathrm{d}N/\mathrm{d}S$) were obtained in `scoup`
where they are calculated using expressions from @spielman2015. The
corresponding maximum likelihood estimates of the ratio ($\omega$) were
obtained using PAML [@yang2007]. The control files used for the execution of
the PAML analyses are available in the `extdata/example/` folder in the
`queeems` package. The following code is consequently necessary to read the
$\mathrm{d}N/\mathrm{d}S$ and the $\omega$ estimates into the current session.
```{r, eval=FALSE}
# ><>< # Read dN/dS values from file
dndsfile <- file.path(datadir, "tut_dnds.csv")
dndsvalues <- read.table(dndsfile, header=TRUE)

# ><>< # Read MLE of non-synonymous/synonymous substitution from file
mlefile <- file.path(datadir, "tut_mles.txt")
mlevalues <- read.table(mlefile, header=TRUE)

# ><>< # Temporary storage of the saturation free sequences.
# ><>< # Data was retrieved from where it is saved and analysed with PAML
message("\nSequences that are rid of saturated sites",
    " may be retrieved from:\n\t", nosatpath, "\n")

# ><>< # Read MLE of substitution rate ratio for saturation-free sequences.
# mle2file <- file.path(datadir, "tut_cleanmle.txt")
# updatedMLE <- read.table(mle2file, header=TRUE)
```

Next, I interrogate the outputs from the analyses. I begin with the entropy
estimates. Theoretically, full saturation occurs when the proability of
observing a base is equal for all bases in the sample space. This corresponds
to the case where the Shannon entropy is equal to $\log(1/n)$, where `n` is
the size (`n`) of the respective sample space. That is, `n = 4` for
nucleotide, `n = 61` for codon and `n = 20` for amino acid bases. The red
dashed lines in the image below indicate the point of theoretical saturation
and the thin lines represent the entropy density for each of the eight
simulated genetic sequence replicate. The same data sets were used for all
the three plots. As expected, nucleotide entropy estimates are potentially
closest to saturation and the codon estimates are furthest. In other words,
`n` can be loosely said to be proportional to the capacity to retain genetic
information.

```{r, eval=FALSE}
# ><>< # Generate distributions of entropy estimates
par(fig=c(0, 0.36, 0, 1), mar=c(4.2,4.2,0.2,0))
    plot(0, 0, col="white", bty="n", xlab="", ylab="",
        xlim=c(0,1.4), ylim=c(0.04,1.2), xaxt="n", yaxt="n")
    axis(1, tck=-0.035, lwd=2, padj=-.5, cex.axis=1.2)
    axis(2, las=1, tck=-0.035, lwd=2, cex.axis=1.2, hadj=0.7)
    a1 <- apply(nucent, 2, function(i) lines(density(i), lwd=.5, col="grey60"))
    lines(rep(-log(1/4),2), c(-0.1,1.3), lwd=2, lty=2, col="tomato")
    mtext("Nuc. Entropy", 1, 2.2, cex=1.3)
    mtext("Density", 2, 2.4, cex=1.3)
    box(lwd=2)
par(fig=c(0.36, 0.68, 0, 1), mar=c(4.2,0,0.2,0), new=TRUE)
    plot(0, 0, col="white", bty="n", xlab="", ylab="",
        xlim=c(0,4.2), ylim=c(0.04,1.2), xaxt="n", yaxt="n")
    axis(1, tck=-0.035, lwd=2, padj=-.5, cex.axis=1.2)
    a1 <- apply(codent, 2, function(i) lines(density(i), lwd=.5, col="grey60"))
    lines(rep(-log(1/61),2), c(-0.1,1.3), lwd=2, lty=2, col="tomato")
    mtext("Codon Entropy", 1, 2.2, cex=1.3)
    box(lwd=2)
par(fig=c(0.68, 1, 0, 1), mar=c(4.2,0,0.2,0.2), new=TRUE)
    plot(0, 0, col="white", bty="n", xlab="", ylab="",
        xlim=c(0,3.1), ylim=c(0.04,1.2), xaxt="n", yaxt="n")
    axis(1, tck=-0.035, lwd=2, padj=-.5, cex.axis=1.2)
    a1 <- apply(aasent, 2, function(i) lines(density(i), lwd=.5, col="grey60"))
    lines(rep(-log(1/20),2), c(-0.1,1.3), lwd=2, lty=2, col="tomato")
    mtext("Amino Acid Entropy", 1, 2.2, cex=1.3)
    box(lwd=2)
```

The more useful practical saturation thresholds should be values less than
the theoretical values [@xia2002]. Figure A below shows estimates of
$\log(\omega)$ (filled circles) and $\log(\mathrm{d}N/\mathrm{d}S)$
(open squares) for each of the data replicates. The table below contains
summaries of the saturation analyses of the data sets. It can be observed
that half (= 4/8) of the $\omega$ estimates are significantly inflated.
In line with the identical setting used to generate the sequences, `queeems`
detected saturated sites for all the replicates. The package also correctly
inferred #### `???%` (= `???/4`) of the sequences that produced the inflated
$\omega$ estimates to be saturated for purposes of natural selection analyses.
#### Figure B shows comparison of $\omega$ estimates from the saturated and
the non-saturated subsets for each of the eight replicates. The comparison
shows that `queeems` can only improve inferences of the magnitude of natural
selection effect. The summary table also contains the maximum substitution
saturation indices ($\mathrm{I}_{ss}$) inferred with DAMBE. Given that none
of the estimated DAMBE indices are greater than the minimum critical
$\mathrm{I}_{ss}$ (because, $\min(\mathrm{I}_{ss.c}) = 0.81$), the
simulated data sets may still be able to reproduce the balanced phylogeny
they were simulated from. This emphasises the need for a genetic saturation
assessment tool like `queeems` that is designed to improve natural selection
inferences.
```{r, eval=TRUE}
a <- 1
plot(0,0)
```

---
|DAMBE                |        |queeems              |        |
|:--------------------|:------:|:--------------------|:------:|
|Invariant sites prop.|0.0000  |Invariant sites prop.|a       |
|Mean H               |0.4457  |Saturated sires      |a       |
|Std. error of H      |0.0148  |Overall Bayes factor |a       |
|Iss                  |0.2698  |Overall decision     |Not sat.|
|Iss.c (symmetric)    |0.8093  |                     |        |
|Iss.c (asymmetric)   |0.7095  |                     |        |
|Overall decision     |Not sat.|                     |        |
Table: Comparative saturation analyses outputs that demonstrate the
effectiveness of the inference functions in `queeems`.

# Conclusion
I demonstrated that, compared to the popular existing software for analysing
molecular saturation, `queeems` avails the opportunity to allow researchers
more easily identify the codon sites contributing to overall protein sequence
saturation. I further showed that such knowledge may lead to improved estimate
of the magnitude of the natural selection pressure obtained from such protein
data. The choice of the weight parameters (`size = 6`, `proportion < 1 =
0.4`, `maximum = 7`) and the tolerance for saturation sites (`piSS = 0.40`)
that are used for the `queeems` analyses presented here is arbitrary. A study
about the sensitivity of the parameters to the inferences obtained is ongoing
and intended to be submitted for publication upon completion.

# Citation
A more appropriate citation will be provided for the package in due course.
In the meantime, kindly cite the package as, <em>Sadiq, H. 2026. "queeems:
Quantify the Extent of Evolutionary Evidence in Molecular Sequences".
Bioconductor R Package.</em>

# References

<div id="refs"></div>

# Session info

The output of `sessionInfo()` from the computer where this file is generated
is provided below.

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
