---
title: "queeems: Quantify the Extent of Evolutionary Evidence in Molecular
Sequences"
author:
  - name: Hassan Sadiq
    affiliation:
    - Department of Statistics and Actuarial Science,
        Stellenbosch University, South Africa
    email: hassan.t.sadiq@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
package: queeems
bibliography: queeems.bib
vignette: |
    %\VignetteIndexEntry{queeems Tutorial}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
    BiocStyle::markdown()
```

# Introduction
As with every other statistical investigation, accuracy of inferences from
analyses of genetic sequences depends on the data quality. Assessing the
quality of molecular sequences for biological studies requires quantifying
the degree of evolutionary signature retained in the data. Such an assessment
is popularly done using the saturation (*H*) index introduced by @xia2002.
The index is an extension of @shannon1948 information entropy and it is not
always robust to the evolutionary tree that must have generated the data. In
general, a single H index is estimated for an alignment of molecular sequences.
The significance of the evolutionary information retained in the data is then
assessed with the Student's t hypothesis test that compares the estimated
index to a pre-determined threshold. Xia's H index is an invaluable
contribution to computational evolutionary biology studies. It is however
best suited for phylogeny reconstruction inquiries. `queeems` is a R
[@cran2025] package hosted on the Bioconductor [@bioc2004] platform that
contains Bayesian functions that can be used to assess the quality of
genetic data beyond just for phylogeny reconstruction purposes. It can also
be used to assess how suitable molecular sequences are for broader
evolutionary analyses, including natural selection investigations. In this
practical application of some of the primary functions in the package, I use
simulated sequences to demonstrate some utilities of the package.


# Installation
Use the following code to install `queeems` from the Bioconductor platform.
```{r, eval=FALSE}
if(!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("queeems")
```
The developmental version of the package that contains the most recent
updates, is available under the `devel` tab of the Bioconductor platform
and can be installed by running the following code, just before the final
install line above: `BiocManager::install(version="devel")`.

# Data
The molecular sequences analysed herein are generated with the `scoup`
[@sadiq2026] package that allows genetic sequences to be simulated following
a simplified Ornstein-Uhlenbeck (OU) process. A R file that contains the
simulation code and a folder that contains the simulated data are available
in the `inst/extdata/example/` folder within the `queeems` package.

Genetic sequence alignments were generated on a balanced bifurcating
evolutionary tree that has 64 leaves. For each data simulated, the length
of the internal and the terminal branches were all kept equal to `0.10` for
each execution. The *variance of synonymous selection coefficients* was
specified as, $\sigma_{s}^{2}=0.01$ to ensure a positive natural selection
effect is imposed [@spielman2015]. Eight data replicates were generated.
Other necessary `scoup` simulation parameters specified include: *number of
codon sites = 250*, *OU reversion rate = 0.00*, *OU asymptotic variance =
0.00*, *OU asymptotic mean = 0.00*, *variance of non-synonymous selection
coefficients = 0.50*.

# Simulation Studies
Saturation analyses of the simulated sequences are demonstrated next with
functions from the `queeems` package. The code chunk below contains commands
that analyse each of the simulated data in turn. The statistics collected for
each of the eight replicates  include: (a.) site-wise Shannon entropy
estimates with respect to nucleotide, codon and amino acid bases using the
`queeems::molentropy` function and (b.) measure of the significance of the
evolutionary information contained in the sequences, in terms of Bayes factors
that were generated by codon site and as a single overall estimate for each
genetic sequence data using the `queeems::seqSaturation` function.
```{r, eval=FALSE}
# ><>< # Load package into current R session
# library("queeems")

# ><>< # Retrieve path to simulated tutorial data folder
datadir <- system.file("extdata/example/", package="queeems")

# ><>< # Read merged molecular sequences data as text
seqsname <- file.path(datadir, "tut_seqs.txt")
seqstext <- readLines(seqsname)

# ><>< # Extract information about number of simulated replicates
sbreaks <- which(seqstext == "  64  750")
nreps <- length(sbreaks)

# ><>< # Create data frames to save interesting outputs
homoBF <- vector("numeric", nreps)
cnames <- paste0("r", seq(1,nreps))
nucent <- matrix(NA, 750, nreps, dimnames=list(NULL,cnames))
codent <- matrix(NA, 250, nreps, dimnames=list(NULL,cnames))
aasent <- matrix(NA, 250, nreps, dimnames=list(NULL,cnames))
sbayes <- matrix(NA, 250, nreps, dimnames=list(NULL,cnames))

# ><>< # Storage for saturation-free data to allow subsequent analyses
nosatpath <- file.path(tempdir(), "satfree.txt")

# ><>< # Set preferred Bayes factor threshold
bflimit <- 3

# ><>< # Set preferred saturated sites tolerance
ssprop <- 0.01

# ><>< # Adhoc function to extract saturation-free data
printsite <- function(b0, sqarray, ipath=nosatpath){
    ntag <- rownames(sqarray)[b0]
    sqline <- paste(sqarray[b0,], collapse="")
    fulltxt <- paste0(">", ntag, "\n", sqline)
    write.table(fulltxt, ipath, TRUE, FALSE, col.name=FALSE, row.names=FALSE)
    return(0)
}

# ><>< # Analyse data replicates captured from merged data file
for(a0 in seq(1,nreps)){

    # ><>< # Extract and temporarily save a sequence replicate
    init <- sbreaks[a0] + 1
    halt <- sbreaks[a0] + 128    # >< # Each replicate is from 64 leaves
    seqrep <- seqstext[seq(init,halt)]
    tempfile <- file.path(tempdir(), "iseqs.fasta")
    write.table(seqrep, file=tempfile, append=FALSE,
        quote=FALSE, row.names=FALSE, col.names=FALSE)

    # ><>< # Execute saturation analysis
    satest <- seqSaturation(tempfile, "A", 6, 0.4, 7, bflimit, ssprop)
    homoBF[a0] <- as.numeric( summary(satest)["fullBF",])
    sbayes[,a0] <- BFs(satest)

    # ><>< # Extract and save saturation-free sequence locations
    nullsites <- which(BFs(satest) > bflimit)
    sieved <- seqfilter(tempfile, nullsites, "codon")
    imatrix <- as.matrix(sieved)
    itext <- paste0("  ", nrow(imatrix), "  ", ncol(imatrix))
    write.table(itext,nosatpath,a0!=1,FALSE,row.names=FALSE,col.names=FALSE)
    pholder <- vapply(seq(1,nrow(imatrix)), printsite, imatrix, FUN.VALUE=0)
    write.table("\n",nosatpath,TRUE,FALSE,row.names=FALSE,col.names=FALSE)

    # ><>< # Execute codon entropy analysis
    entcdn <- molentropy(tempfile, "codon", "shannon")
    codent[,a0] <- sitentropies(entcdn)

    # ><>< # Process nucleotide entropy analysis
    entnuc <- molentropy(tempfile, "dna", "shannon")
    nucent[,a0] <- sitentropies(entnuc)

    # ><>< # Transform nucleotide to amino acid sequences
    dnaformat <- Biostrings::readDNAStringSet(tempfile)
    aaformat <- Biostrings::translate(dnaformat)
    Biostrings::writeXStringSet(aaformat, tempfile)

    # ><>< # Implement amino acid entropy analysis
    entamino <- molentropy(tempfile, "aa", "shannon")
    aasent[,a0] <- sitentropies(entamino)

    # ><>< # Provide analysis progress update
    message("Completed analyses of data replicate ", a0)
}
```

For completeness and and an unbiased quality assessment of `queeems`, a few
other statistics about the simulated sequences needed to be obtained using
external resources. This include: (a.) analytical estimates of the ratio of
non-synonymous to synonymous substitution rates ($\mathrm{d}N/\mathrm{d}S$).
These were obtained together with the sequences from `scoup`, where they are
calculated using expressions from @spielman2015. (b.) the maximum likelihood
estimates of the same substitution rates ratio ($\omega$) that were obtained
using the M0 model [@goldman1994; @muse1994] in PAML [@yang2007]. The control
files used for the execution of the PAML analyses are available in the
`extdata/example/` folder within the `queeems` package. The following chunk
of code is necessary to read the external statistics into the current session
for analysis.
```{r, eval=FALSE}
# ><>< # Read dN/dS values from file
dndsfile <- file.path(datadir, "tut_dnds.csv")
dndsvalues <- read.table(dndsfile, header=TRUE)

# ><>< # Read MLE of non-synonymous/synonymous substitution from file
mlefile <- file.path(datadir, "tut_mles.txt")
mlevalues <- read.table(mlefile, header=TRUE)

# ><>< # Temporary storage of the saturation free sequences.
# ><>< # Data was retrieved from where it is saved and analysed with PAML
message("\nSequences that are rid of saturated sites",
    " may be retrieved from:\n\t", nosatpath, "\n")

# ><>< # Read MLE of substitution rate ratio for saturation-free sequences.
mle2file <- file.path(datadir, "tut_cleanmle.txt")
updatedMLE <- read.table(mle2file, header=TRUE)
```

Next, I interrogate the different outputs for the purpose of making a case
for the relevance of `queeems`. I begin with the entropy estimates returned
by the `queeems::molentropy` function. Theoretically, full saturation occurs
when the probability of observing a base is equal for all bases in the sample
space. This happens when the Shannon entropy is equal to $\log(1/n)$, where
`n` is the size of the respective sample space. That is, `n = 4` for
nucleotide, `n = 61` for codon and `n = 20` for amino acid bases. The red
dashed lines in the image below indicate the point of theoretical saturation
and the thin lines represent the entropy density for each of the eight
simulated genetic sequence replicates. The same data sets were used for all
the three plots. As expected, nucleotide entropy estimates are potentially
closest to saturation and the codon estimates are furthest. In other words,
`n` can be loosely said to be proportional to the capacity to retain genetic
information. The entropy results presented here are functions of observed
base frequencies. Other measures of entropy are accessible in `queeems`.
These include: the `cncsentropy` and the `codondifferindex` functions. The
package also allows users to opt for `Renyi` entropy calculation method. 
Reader may consult the help pages of the respective functions for further
details.

```{r, eval=FALSE}
# ><>< # Generate distributions of entropy estimates
par(fig=c(0, 0.36, 0, 1), mar=c(4.2,4.2,0.2,0))
    plot(0, 0, col="white", bty="n", xlab="", ylab="",
        xlim=c(0,1.4), ylim=c(0.04,1.2), xaxt="n", yaxt="n")
    axis(1, tck=-0.035, lwd=2, padj=-.5, cex.axis=1.2)
    axis(2, las=1, tck=-0.035, lwd=2, cex.axis=1.2, hadj=0.7)
    a1 <- apply(nucent, 2, function(i) lines(density(i), lwd=.5, col="grey60"))
    lines(rep(-log(1/4),2), c(-0.1,1.3), lwd=2, lty=2, col="tomato")
    mtext("Nuc. Entropy", 1, 2.2, cex=1.3)
    mtext("Density", 2, 2.4, cex=1.3)
    box(lwd=2)
par(fig=c(0.36, 0.68, 0, 1), mar=c(4.2,0,0.2,0), new=TRUE)
    plot(0, 0, col="white", bty="n", xlab="", ylab="",
        xlim=c(0,4.2), ylim=c(0.04,1.2), xaxt="n", yaxt="n")
    axis(1, tck=-0.035, lwd=2, padj=-.5, cex.axis=1.2)
    a1 <- apply(codent, 2, function(i) lines(density(i), lwd=.5, col="grey60"))
    lines(rep(-log(1/61),2), c(-0.1,1.3), lwd=2, lty=2, col="tomato")
    mtext("Codon Entropy", 1, 2.2, cex=1.3)
    box(lwd=2)
par(fig=c(0.68, 1, 0, 1), mar=c(4.2,0,0.2,0.2), new=TRUE)
    plot(0, 0, col="white", bty="n", xlab="", ylab="",
        xlim=c(0,3.1), ylim=c(0.04,1.2), xaxt="n", yaxt="n")
    axis(1, tck=-0.035, lwd=2, padj=-.5, cex.axis=1.2)
    a1 <- apply(aasent, 2, function(i) lines(density(i), lwd=.5, col="grey60"))
    lines(rep(-log(1/20),2), c(-0.1,1.3), lwd=2, lty=2, col="tomato")
    mtext("Amino Acid Entropy", 1, 2.2, cex=1.3)
    box(lwd=2)
```

Genetic sequences are expected to be almost useless for evolutionary
analyses before they reach theoretical saturation point [@xia2002]. The
objective when developing an effective saturation measure is therefore
to uncover whether a given set of homologous genetic sequences have lost
practical relevance. The primary technique in `queeems` is to identify
saturated base sites and sequence set with Bayes factors that are
estimated by comparing codon to nucleotide entropies. The following
outputs demonstrate the potency of this identification approach.
```{r, eval=FALSE}
# ><>< # Plot estimates of the magnitude of natural seletion effect
par(fig=c(0, 0.52, 0, 1), mar=c(4.2,4.2,0.2,0))
    plot(0, 0, col="white", bty="n", xlab="", ylab="",
        xlim=c(.5,8.5), ylim=c(1.5,7), xaxt="n", yaxt="n")
    axis(1, tck=-0.035, lwd=2, padj=-.5, cex.axis=1.2)
    axis(2, las=1, tck=-0.035, lwd=2, cex.axis=1.2, hadj=0.7)
    mtext(expression(paste("Log (",omega,", dN/dS)")), 2, 2.4, cex=1.3)
    mtext("Simulated Data", 1, 2.2, cex=1.3)
    points(log(c(as.matrix(mlevalues))), cex=3, pch=16)
    points(log(c(as.matrix(dndsvalues))), cex=3, pch=0, lwd=2)
    box(lwd=2)
par(fig=c(0.52, 1, 0, 1), mar=c(4.2,0.1,0.2,0.2), new=TRUE)
    plot(0, 0, col="white", bty="n", xlab="", ylab="",
        xlim=c(.5,8.5), ylim=c(1.5,7), xaxt="n", yaxt="n")
    axis(1, tck=-0.035, lwd=2, padj=-.5, cex.axis=1.2)
    mtext("Simulated Data", 1, 2.2, cex=1.3)
    points(log(c(as.matrix(mlevalues))), cex=3, pch=16)
    points(log(c(as.matrix(dndsvalues))), cex=3, pch=0, lwd=2)
    box(lwd=2)

# ><>< # Tabulate interesting summaries
print(homoBF)   # Homogeneous/Overall Bayes Factor for each replicate

satsites <- apply(sbayes, 2, function(i) sum(i > bflimit))
print(satsites) # Number of saturated sites identified per replicate
```

Figure A above shows estimates of
$\log(\omega)$ (filled circles) and $\log(\mathrm{d}N/\mathrm{d}S)$
(open squares) for each of the data replicates. The table below contains
summaries of the saturation analyses of the data sets. It can be observed
that half (= 4/8) of the $\omega$ estimates are significantly inflated.
In line with the identical setting used to generate the sequences, `queeems`
detected saturated sites for all the replicates. The package also correctly
inferred #### `???%` (= `???/4`) of the sequences that produced the inflated
$\omega$ estimates to be saturated for purposes of natural selection analyses.
#### Figure B shows comparison of $\omega$ estimates from the saturated and
the non-saturated subsets for each of the eight replicates. The comparison
shows that `queeems` can only improve inferences of the magnitude of natural
selection effect. The summary table also contains the maximum substitution
saturation indices ($\mathrm{I}_{ss}$) inferred with DAMBE [@xia2018]. Given
that none of the estimated DAMBE indices are greater than the minimum
critical $\mathrm{I}_{ss}$ (because, $\min(\mathrm{I}_{ss.c}) = 0.81$), the
simulated data sets may still be able to reproduce the balanced phylogeny
they were simulated from. This emphasises the need for a genetic saturation
assessment tool like `queeems` that is designed to improve natural selection
inferences.

---
|  |DAMBE     |            |queeems|          |
|:-|----------|:-----------|-------|----------|
|  |*Max. Iss*|*Max. Iss.c*|*BF*   |*SS Count*|
|R1|  0.508   |   0.810    | 2.6938|  5 / 250 |
|R2|  0.487   |   0.810    |48.1838|  8 / 250 |
|R3|  0.515   |   0.810    | 0.9724|  3 / 250 |
|R4|  0.503   |   0.810    | 0.6450|  2 / 250 |
|R5|  0.493   |   0.810    |15.7384|  7 / 250 |
|R6|  0.522   |   0.810    |15.7384|  7 / 250 |
|R7|  0.502   |   0.810    | 1.4660|  4 / 250 |
|R8|  0.508   |   0.810    |48.1838|  8 / 250 |
Table: Comparative saturation analyses outputs that demonstrate the
effectiveness of the inference functions in `queeems`. DAMBE outputs
were obtained from @xia2018. R$\star$ = simulated data replicate $\star$.
Each replicate comprised 250 codon sites and 64 sequences simulated on a
balanced evolutionary tree using `scoup` [@sadiq2026]. $\mathrm{I}_{ss}$
= substitution saturation index and $\mathrm{I}_{ss.c}$ = critical
$\mathrm{I}_{ss}$. DAMBE returns multiple indices for data sets that
contain more than 16 sequences. The reported indices are the maximum
estimates. SS Count = number of saturated sites detected. BF = overall
Bayes factor.

# Conclusion
I demonstrated that, compared to the popular existing software for analysing
molecular saturation, `queeems` avails increased flexibility that enables
researchers to more easily identify the codon sites that are likely
contributing to overall genetic sequence saturation. I further showed that
such knowledge may lead to improved estimate of the magnitude of the natural
selection pressure obtained from such molecular data. The choice of the
weight parameters, Bayes factor threshold and the tolerance for saturation
sites that are used for the `queeems` analyses presented here is arbitrary.
A study about the sensitivity of the parameters is nearing completion.

# Citation
A more appropriate citation will be provided for the package in due course.
In the meantime, kindly cite the package as, <em>Sadiq, H. 2026. "queeems:
Quantify the Extent of Evolutionary Evidence in Molecular Sequences".
Bioconductor R Package.</em>

# References

<div id="refs"></div>

# Session info

The output of `sessionInfo()` from the computer where this file is generated
is provided below.

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
