---
title: "Quantify the Extent of Evolutionary Evidence in Molecular Sequences"
author:
- name: Hassan Sadiq
  affiliation: Department of Statistics and Actuarial Science,
    Stellenbosch University, South Africa
output:
    BiocStyle::html_document:
        toc_float: true
package: queeems
bibliography: citation.bib
abstract: |
    Sewall Wright aaa
vignette: |
    %\VignetteIndexEntry{queeems Tutorial}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
    BiocStyle::markdown()
```

# Introduction
Accuracy of inferences from analyses of molecular data, as with every other
statistical investigation, depends on the data quality. Assessing the
quality of molecular sequences for biological studies requires quantifying
the degree of evolutionary signature retained in the data. Such an assessment
is popularly done using a saturation (H) index that is inspired by information
entropy. A single H index is estimated for an alignment of protein sequences
and the significance of the evolutionary information retained is assessed
with the Student's t hypothesis test. The H index is an invaluable
contribution to computational evolutionary biology studies. Like many other
existing measures of genetic data quality, the H index is most suited for
phylogeny reconstruction inquiries.


# Installation
Use the following code to install `queeems` from the Bioconductor platform.
```{r, eval=FALSE}
if(!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("queeems")
```
The developmental version of the package that contains the most recent
updates, is available under the `devel` tab of the Bioconductor platform.

# Sample Code
Three primary evolutionary algorithms .... Example of `R`
[@cran2024] ... package.

## Ornstein-Uhlenbeck Sensitivity
```{r, eval=TRUE}
# Make package accessible in R session
library(scoup)

# Number of extant taxa
leaves <- 8

# Number of codon sites
sSize <- 15

# Number of data replications for each parameter combination
sims <- 1

# OU reversion parameter (Theta) value
eThta <- c(0.01)

# OU asymptotic variance value
eVary <- c(0.0001)

# OU landscape shift parameters
hbrunoStat <- hbInput(c(vNvS=1, nsynVar=0.01))

# Sequence alignment size information
seqStat <- seqDetails(c(nsite=sSize, ntaxa=leaves))

# Iterate over all listed OU variance values
for(g in seq(1,length(eVary))){

    # Iterate over all listed OU reversion parameter values
    for(h in seq(1,length(eThta))){

        # Create appropriate simulation function ("ou") object
        adaptStat <- ouInput(c(eVar=eVary[g],Theta=eThta[h]))

        # Iterate over the specified number of replicates
        for(i in seq(1,sims)){

            # Execute simulation
            simData <- alignsim(adaptStat, seqStat, hbrunoStat, NULL)
        }
    }
}
# Print simulated alignment
# seqCOL(simData)
```

# Conclusion
....

# Citation
A more appropriate citation will be provided for the package in due course.
In the meantime, kindly cite the package as, <em>Sadiq, H. 2026. "queeems:
Quantify the Extent of Evolutionary Evidence in Molecular Sequences".
Bioconductor R Package. doi:10.18129/B9.bioc.queeems.</em>

# References

<div id="refs"></div>

# Appendix: Sample Data Analyses (Non-R) Code
Example `CODEML` script executed in `PAML` to infer single alignment-wide
$\omega$ estimates for some of the data sets analysed above.
```{bash, eval=FALSE}
    seqfile   = seq.nex
    treefile  = seq.tre
    outfile   = seq.out
    noisy     = 0
    verbose   = 0
    seqtype   = 1
    ndata     = 1
    icode     = 0
    cleandata = 0
    model     = 0
    NSsites   = 0
    CodonFreq = 3
    estFreq   = 0
    clock     = 0
    fix_omega = 0
    omega     = 1e-05
```

# Session info

The output of `sessionInfo()` from the computer where this file is generated
is provided below.

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
